---
title: "SpatialExperiment and SpaceTrooper"
subtitle: "Unified infrastructure and quality control workflows for spatial omics data"
author: "Dario Righelli"
date: today
date-format: "D MMMM YYYY"
output-file: index.html
execute:
  echo: true
  warning: false
  message: false
engine: knitr
editor: source
---

<!--
WORKSHOP STRUCTURE (~180 min)

00-10  Spatial omics in Bioconductor + OSTA
10-25  SpatialExperiment from scratch: synthetic data
25-45  Real data: SPE infrastructure + ggspavis + molecule assays
45-50  Beyond SpatialExperiment
50-55  IO ecosystem
55-65  Break
65-85  SpaceTrooper: spatial-aware quality control
85-125 SpaceTrooper hands-on
125-155 QScore, spatial diagnostics & interpretation
155-180 Exercises / discussion / outlook

-->

::: {.hero}
<div class="eyebrow">BIOCONDUCTOR WORKSHOP</div>

# SpatialExperiment and SpaceTrooper

<div class="hero-subtitle">Unified infrastructure and quality control workflows for spatial omics data</div>

<div class="hero-meta">SpatialExperiment · OSTA · SpaceTrooper</div>
:::

## Learning goals

By the end of this workshop, you should be able to:

- understand how Bioconductor represents spatial omics experiments;
- inspect and manipulate a `SpatialExperiment` object;
- recognize how the same infrastructure can accommodate very different spatial technologies;
- understand how real sequencing- and imaging-based datasets reuse the same `SpatialExperiment` infrastructure;
- visualize spot/cell- and molecule-level spatial information with `ggspavis`;
- understand how a `BumpyMatrix` can store molecule-level information inside a `SpatialExperiment` assay;
- understand where `SpatialFeatureExperiment`, `MoleculeExperiment`, and spatial IO packages fit in the ecosystem;
- apply spatial-aware quality control using `SpaceTrooper`.

# Spatial omics in Bioconductor

<!-- ~10 min -->

## Spatial omics: one name, different technologies

Spatial transcriptomics technologies can be broadly separated into two families:

### Sequencing-based

Examples include:

- Visium / Visium HD
- Slide-seq and related bead-based approaches

Typical observational units are **spots or bins**.

### Imaging-based

Examples include:

- Xenium
- CosMx
- MERSCOPE

Typical observational units are **segmented cells**, while some platforms also provide
molecule-level coordinates and cell or nuclear boundaries.

The important point for this workshop is not the vendor or platform.

::: {.callout-note title="Key idea"}
**Different technologies generate different spatial information, but we still want common data structures and common analysis interfaces.**
:::

## OSTA: a map of the Bioconductor spatial ecosystem

**OSTA — Orchestrating Spatial Transcriptomics Analysis with Bioconductor** provides
an ecosystem-level view of spatial transcriptomics workflows in Bioconductor.

It covers both major technology families and connects:

- data infrastructure;
- data import;
- quality control;
- preprocessing;
- visualization;
- downstream statistical analysis.

For this workshop, we use OSTA as a **map of the ecosystem**, while focusing on
the infrastructure that makes interoperability possible.

## From an ecosystem to a common representation

A spatial experiment may include:

- a feature-by-observation assay matrix;
- feature metadata;
- observation metadata;
- spatial coordinates;
- images and image metadata;
- technology-specific spatial information.

The question is:

::: {.callout-important title="The infrastructure question"}
**How can analysis packages interact with all of these experiments without implementing a different interface for every platform?**
:::

# SpatialExperiment: a flexible infrastructure

<!-- ~20 min -->

## Building on Bioconductor infrastructure

`SpatialExperiment` extends the existing Bioconductor experiment hierarchy:

```text
SummarizedExperiment
        |
        v
SingleCellExperiment
        |
        v
SpatialExperiment
```

This means that a spatial experiment keeps the familiar Bioconductor components:

```r
assays(spe)
rowData(spe)
colData(spe)
```

and adds spatially relevant information:

```r
spatialCoords(spe)
imgData(spe)
```

## A SpatialExperiment is not an analysis workflow

The role of `SpatialExperiment` is **representation**, not analysis.

It does not prescribe:

- normalization;
- clustering;
- spatial statistics;
- segmentation;
- differential expression;
- quality control.

Instead, it provides a common object on which downstream methods can operate.

::: {.callout-tip title="Take-home message"}
**SpatialExperiment describes what the experiment contains, not how it must be analyzed.**
:::

## One class, different technologies

The same class can represent experiments with very different observational units.

| | Sequencing-based | Imaging-based |
|---|---|---|
| Typical unit | spot / bin | cell |
| Assay | gene counts | gene counts |
| Coordinates | spot/bin coordinates | cell centroids |
| Image | often present | often present |
| Segmentation | usually not central | often important |
| Molecules | usually not available | may be available |

The common denominator is deliberately small.

This is what makes the infrastructure reusable.

# Hands-on I: build the abstraction from scratch

## Start with deliberately simple data

Before looking at a real platform, we create two **small teaching objects** in
memory:

1. a spot-based, sequencing-like experiment;
2. a cell-based, imaging-like experiment.

These objects are intentionally synthetic. At this stage we want to isolate the
**data model** from the complexity of vendor files, biological annotation, and
large datasets.

```{r}
#| label: setup-spe
library(SpatialExperiment)
library(S4Vectors)

set.seed(123)
```

## A spot-based experiment

```{r}
#| label: make-spot-spe

nGenes <- 100
nSpots <- 36

spotCounts <- matrix(
    rpois(nGenes * nSpots, lambda=5),
    nrow=nGenes,
    ncol=nSpots,
    dimnames=list(
        paste0("gene_", seq_len(nGenes)),
        paste0("spot_", seq_len(nSpots))
    )
)

spotGrid <- expand.grid(
    x=seq_len(6),
    y=seq_len(6)
)

speSeq <- SpatialExperiment(
    assays=list(counts=spotCounts),
    rowData=DataFrame(
        gene_id=rownames(spotCounts)
    ),
    colData=DataFrame(
        technology=rep("sequencing-based", nSpots),
        unit=rep("spot", nSpots)
    ),
    spatialCoords=as.matrix(spotGrid)
)

speSeq
```

## A cell-based imaging experiment

```{r}
#| label: make-imaging-spe

nCells <- 120

cellCounts <- matrix(
    rpois(nGenes * nCells, lambda=2),
    nrow=nGenes,
    ncol=nCells,
    dimnames=list(
        paste0("gene_", seq_len(nGenes)),
        paste0("cell_", seq_len(nCells))
    )
)

cellCoords <- cbind(
    x=runif(nCells, 0, 1000),
    y=runif(nCells, 0, 1000)
)

speImg <- SpatialExperiment(
    assays=list(counts=cellCounts),
    rowData=DataFrame(
        gene_id=rownames(cellCounts)
    ),
    colData=DataFrame(
        technology=rep("imaging-based", nCells),
        unit=rep("cell", nCells),
        cell_area=rgamma(nCells, shape=10, scale=8),
        nucleus_area=rgamma(nCells, shape=8, scale=4)
    ),
    spatialCoords=cellCoords
)

speImg
```

## Same interface, different spatial units

Run the **same commands** on both objects.

```{r}
#| label: inspect-synthetic-spe

dim(speSeq)
dim(speImg)

assayNames(speSeq)
assayNames(speImg)

head(colData(speSeq))
head(colData(speImg))

head(rowData(speSeq))
head(rowData(speImg))

head(spatialCoords(speSeq))
head(spatialCoords(speImg))
```

A minimal plot is already possible using only the common spatial interface.

```{r}
#| label: plot-synthetic-spe
#| fig-width: 7
#| fig-height: 4.5

plot(
    spatialCoords(speSeq),
    asp=1,
    pch=16,
    xlab="x",
    ylab="y",
    main="Synthetic spot-based experiment"
)
```

::: {.callout-tip title="Take-home message"}
**The observational unit changes; the Bioconductor interface does not.**
:::

# Hands-on II: the same infrastructure on real data

The synthetic objects showed the abstraction. Now we ask whether it survives
contact with real data -- and whether downstream packages can make use of it.

We use example datasets already distributed in Bioconductor and visualize them
with `ggspavis`.

```{r}
#| label: setup-real-spe
library(STexampleData)
library(ggspavis)
library(BumpyMatrix)
```

## Real sequencing-based data: Visium

Load a real 10x Genomics Visium section from `STexampleData`.

```{r}
#| label: load-real-visium
speVisium <- Visium_humanDLPFC()

speVisium
```

The object is larger and biologically meaningful, but the interface is the one
we have already seen.

```{r}
#| label: inspect-real-visium
class(speVisium)
dim(speVisium)
assayNames(speVisium)

head(rowData(speVisium))
head(colData(speVisium))
head(spatialCoords(speVisium))
```

### Visualization through the infrastructure

`ggspavis` does not need to know how the original Space Ranger files were
organized. It receives a `SpatialExperiment` and uses the standardized
information stored in the object.

You can plot the spots coordinates with the ground truth annotation.

```{r}
#| label: ggspavis-visium
#| fig-width: 7
#| fig-height: 6

plotCoords(
    speVisium,
    point_size = 0.5,
    annotate="ground_truth"
)
```

Or you can plot the visium experiment with a `colData` or `rowData` annotation and highlight only the in-tissue spots.

```{r}
plotVisium(speVisium, annotate="ground_truth", highlight="in_tissue", 
                 legend_position="none", point_size=0.5)
```
The point is not the particular annotation. The important step is:

```text
Visium files -> SpatialExperiment -> ggspavis
```

Once the data are represented by the common class, visualization can operate on
the infrastructure rather than on a vendor-specific file layout.

## Real imaging-based data: seqFISH

Imaging-based technologies can also be represented at the cell level by a
`SpatialExperiment`.

```{r}
#| label: load-real-seqfish
speSeqFISH <- seqFISH_mouseEmbryo()

speSeqFISH
```

Again, start from the same interface.

```{r}
#| label: inspect-real-seqfish
class(speSeqFISH)
dim(speSeqFISH)
assayNames(speSeqFISH)

head(colData(speSeqFISH))
head(spatialCoords(speSeqFISH))
```

### Cell-level representation

At the aggregated cell level, the imaging experiment looks familiar: genes by
cells plus spatial coordinates.

```{r}
#| label: ggspavis-seqfish-cells
#| fig-width: 7
#| fig-height: 6

plotCoords(
    speSeqFISH,
    in_tissue=NULL,
    y_reverse=FALSE
)
```

## Imaging data can retain individual molecule coordinates

This is where an imaging-based experiment can contain information that does not
fit naturally into a simple numeric gene-by-cell matrix.

`SpatialExperiment` defines a standardized assay named `molecules`. This assay
is usually represented by a `BumpyMatrix`: every gene-by-cell entry can contain
a variable number of molecule observations, for example their `x` and `y`
coordinates.

```{r}
#| label: inspect-molecules-assay
assayNames(speSeqFISH)

class(molecules(speSeqFISH))
dim(molecules(speSeqFISH))

molecules(speSeqFISH)[1:2, 1:2]
```

Conceptually:

```text
                     cell_1             cell_2
Gene A       (x,y), (x,y), ...       (x,y), ...
Gene B              (x,y)          (x,y), (x,y)
Gene C                ...                ...

                  BumpyMatrix
```

This allows a single `SpatialExperiment` to keep both the aggregated counts and
the underlying molecule-resolved information:

```r
spe <- SpatialExperiment(
    assays=list(
        counts=y,
        molecules=mol
    )
)
```

The convenience accessor:

```r
molecules(spe)
```

is equivalent to retrieving the assay named `"molecules"`.

### Visualize individual molecules

The same structure is directly useful downstream. For example, `ggspavis` can
plot the spatial locations of individual transcripts for a selected gene.

```{r}
#| label: ggspavis-molecules
#| fig-width: 7
#| fig-height: 6

plotMolecules(
    speSeqFISH,
    molecule="Sox2"
)
```

::: {.callout-important title="Why this matters"}
A `SpatialExperiment` can therefore provide **both levels of representation**:
cell-level counts and molecule-level coordinates. The common class is not tied
to one spatial resolution.
:::

# Beyond SpatialExperiment

## When points and assays are not enough

The examples above show that `SpatialExperiment` already covers a surprisingly
wide range of use cases:

- spot- or cell-level count matrices;
- spatial coordinates;
- images and image metadata;
- molecule-level coordinates through a `BumpyMatrix` `molecules` assay.

However, some experiments require richer spatial objects and operations than a
matrix plus point coordinates can conveniently express.

### `SpatialFeatureExperiment`

`SpatialFeatureExperiment` extends the `SpatialExperiment` idea with
geometry-aware infrastructure based on simple features (`sf`). This is useful
for objects such as:

- cell and nucleus segmentation polygons;
- tissue boundaries;
- spatial regions and annotations;
- multiple geometries associated with the same experiment.

### `MoleculeExperiment`

`MoleculeExperiment` provides a dedicated representation for molecule-resolved
experiments, with explicit molecule and boundary structures and operations for
summarizing molecules into cell-level representations.

This does **not** mean that molecule-level information cannot live inside a
`SpatialExperiment`: we have just seen the `BumpyMatrix` `molecules` assay.
Instead, the dedicated classes become useful when the spatial representation
itself needs to be richer.

::: {.callout-note title="Beyond SpatialExperiment"}
**Start from a common abstraction; move to richer spatial infrastructure when the experiment requires it.**
:::

# Getting data in: the IO ecosystem

## Separate file formats from experiment representation

Vendor outputs are complicated and change over time.

Bioconductor therefore increasingly separates:

```text
vendor / platform files
          |
          v
      IO layer
          |
          v
Bioconductor experiment objects
```

Examples include:

- `TENxIO` — general infrastructure for reading several 10x Genomics file formats;
- `VisiumIO` — readers for Visium and Visium HD outputs;
- `XeniumIO` — readers for 10x Genomics Xenium outputs;
- `SpatialExperimentIO` — multi-platform readers for imaging-based technologies such as Xenium, CosMx, MERSCOPE, and seqFISH;
- `SpatialFeatureExperiment` — readers that can build geometry-aware `SpatialFeatureExperiment` objects for several spatial platforms.

This separation is important:

::: {.callout-note title="IO principle"}
**Reading files and analyzing experiments are different problems.**
:::

Once the data are represented through standard Bioconductor infrastructure,
downstream methods can focus on the biology and statistics rather than on
vendor-specific parsing.

## Where we are

At this point we have moved through:

```text
OSTA / ecosystem
       |
       v
SpatialExperiment
       |
       +--> sequencing-based experiments
       |
       +--> imaging-based experiments
       |
       +--> richer representations when needed
       |
       +--> platform-specific IO
```

After the break we move from **data infrastructure** to **methodology**.

# Break

::: {.break-card}
<div class="break-big">10 min</div>
<div>Break</div>
:::

# From infrastructure to methods

## Why spatial-aware quality control?

Standard single-cell QC metrics are useful, but imaging-based spatial omics
introduces additional failure modes that are directly connected to image
acquisition, segmentation, morphology, and spatial organization.

The SpaceTrooper paper highlights examples including:

- segmentation errors;
- signal loss;
- elevated background;
- spatial distortions;
- tissue-derived artifacts.

::: {.callout-important title="Why SpaceTrooper?"}
**SpaceTrooper is designed specifically for imaging-based single-cell spatial
omics.** It combines expression-derived and morphological information rather
than relying only on conventional scRNA-seq QC thresholds.
:::

## From fixed thresholds to a data-driven QC score

Conventional QC often looks like this:

```text
metric 1 > threshold
metric 2 < threshold
metric 3 > threshold
        |
        v
    keep / remove
```

This is simple, but each metric is considered separately and thresholds may not
transfer well across tissues, platforms, or experiments.

SpaceTrooper instead follows a data-driven strategy:

```text
expression-derived metrics
          +
morphological metrics
          +
platform-specific effects
          |
          v
 integrated per-cell model
          |
          v
      Quality Score
```

The goal is not to remove the need for biological judgement.

The goal is to provide a **multivariate summary of cell quality** that can be
inspected spatially.

## What contributes to cell quality?

The current SpaceTrooper workflow uses several complementary quantities.

| Metric | What it captures |
|---|---|
| `log2SignalDensity` | transcript signal relative to cell size |
| `Area_um` | cell size / possible segmentation abnormalities |
| `log2Ctrl_total_ratio` | negative-control signal relative to total signal |
| `log2AspectRatio` | abnormal cell elongation / morphology |
| `dist_border` | proximity to a field-of-view border, when relevant (CosMx case) |

For CosMx, aspect ratio and FOV-border distance can jointly capture a
platform-specific **border effect**.

::: {.callout-note title="The main idea"}
A suspicious cell is not necessarily defined by one extreme metric.  
**Several weak QC signals can become informative when considered together.**
:::

## SpaceTrooper in the Bioconductor workflow

This is where the first part of the workshop becomes useful. The analysis layer
does not need to reinvent data representation or vendor-file parsing:

```text
vendor / platform files
          |
          v
 SpatialExperimentIO
 readCosmxSXE()
 readXeniumSXE()
 readMerscopeSXE()
          |
          v
   SpatialExperiment
          |
          +---- SpaceTrooper standardization
          |       updateCosmxSPE()
          |       updateXeniumSPE()
          |
          v
     SpaceTrooper
          |
          +--> per-cell QC metrics
          +--> spatial diagnostics
          +--> Quality Score
          +--> QC flags
```

SpaceTrooper also provides high-level platform readers for a shorter route:

```r
readCosmxSPE()
readXeniumSPE()
readMerfishSPE()
```

::: {.callout-note title="Infrastructure first"}
The reader functions are not separate analysis universes: they converge on a
**`SpatialExperiment`**, which is then enriched with the platform-specific
metadata required by the QC methods.
:::

SpaceTrooper therefore works **on top of `SpatialExperiment`**, while targeting
the specific needs of **imaging-based spatial omics**.

# SpaceTrooper hands-on

<!-- ~40 min -->

## Setup

For the practical we use the **1k-plex CosMx mouse brain dataset** distributed
through `OSTA.data` and used in the OSTA imaging-based QC workflow.

This is a realistically sized experiment: after import it contains approximately **950 targets and 48,556 cells**, distributed across many fields of view. We retrieve cell-level data and boundaries, but skip the molecule-level transcript table because it is not needed for this QC practical.

```{r}
#| label: setup-spacetrooper

library(OSTA.data)
library(SpaceTrooper)
library(SpatialExperiment)
library(SummarizedExperiment)
library(S4Vectors)
library(ggplot2)

set.seed(1713)
```

## Loading infrastructure: from files to `SpatialExperiment`

This dataset is useful because we can make the different infrastructure layers
explicit rather than starting from an already prepared object.

For CosMx, there are two interoperable routes:

```text
Route A: convenience reader

CosMx files
    |
    v
SpaceTrooper::readCosmxSPE()
    |
    v
SpatialExperiment


Route B: composable Bioconductor infrastructure

CosMx files
    |
    v
SpatialExperimentIO::readCosmxSXE()
    |
    v
SpatialExperiment
    |
    v
SpaceTrooper::updateCosmxSPE()
    |
    v
SpaceTrooper-ready SpatialExperiment
```

In the workshop we use **Route A**. This keeps the practical focused on
SpaceTrooper while preserving the common Bioconductor representation:
`readCosmxSPE()` returns a `SpatialExperiment`.

The two routes are interoperable rather than competing alternatives.
`readCosmxSPE()` itself relies on `SpatialExperimentIO::readCosmxSXE()` for the
file import and then performs the SpaceTrooper-specific standardization by aid of the  
`updateCosmxSPE()` function.

::: {.callout-tip title="The same idea extends across platforms"}
SpaceTrooper exposes `readCosmxSPE()`, `readXeniumSPE()`, and
`readMerfishSPE()` as high-level readers. They provide a concise entry point
from vendor outputs to a SpaceTrooper-ready `SpatialExperiment`.
:::

## Retrieve the OSTA CosMx dataset

`OSTA.data_load()` downloads the dataset through the OSTA data infrastructure
and caches the archive. We keep polygons but skip molecule-level coordinates.

```{r}
#| label: download-osta-cosmx

datasetId <- "CosMx1k_MouseBrain2"

archivePath <- OSTA.data_load(
    datasetId,
    pol=TRUE,
    mol=FALSE
)

(dataDir <- file.path(tempdir(), datasetId))
dir.create(dataDir, recursive=TRUE, showWarnings=FALSE)
unzip(archivePath, exdir=dataDir)

dataDir
```

## Step 1 — load CosMx data with SpaceTrooper

`readCosmxSPE()` is the high-level SpaceTrooper entry point. It reads the CosMx
files, constructs a `SpatialExperiment`, and standardizes the metadata required
by the downstream QC workflow.

```{r}
#| label: import-cosmx-spacetrooper

spe <- readCosmxSPE(
    dirName=dataDir
)
spe
```

The returned object is already a `SpatialExperiment`. Target counts and
negative probes remain accessible through the standard Bioconductor interfaces.

```{r}
#| label: inspect-imported-cosmx

class(spe)
dim(spe)
assayNames(spe)
altExpNames(spe)

head(spatialCoords(spe))
head(colData(spe))
head(rowData(spe))
head(metadata(spe)$fov_positions)
metadata(spe)$technology 

```

## Step 2 — attach cell boundaries

The downloaded OSTA archive also contains cell boundaries.
`readAndAddPolygonsToSPE()` adds them to the same experiment object so that
morphological and spatial diagnostics can use them directly.

```{r}
#| label: add-polygons

spe <- readAndAddPolygonsToSPE(spe)
spe$in_tissue <- TRUE

spe
```

::: {.callout-important title="One object, several infrastructure layers"}
We started from platform files, loaded them through SpaceTrooper, and
attached geometry without abandoning the **`SpatialExperiment`** interface.
:::

Before doing any QC, we can see that the colData contains the expected CosMx-specific polygon as sf geometry.

```{r}
#| label: inspect-real-spe
head(colData(spe))
```

## Look at the experiment before computing QC

With tens of thousands of cells and many FOVs, we can now inspect actual spatial
structure rather than a toy example.

```{r}
#| label: plot-cosmx-centroids
#| fig-width: 8
#| fig-height: 6

plotCentroids(spe)
```

## Check the field-of-view geometry

For CosMx, FOV placement is part of the technical structure of the experiment.
A spatial misalignment between FOV coordinates and cell centroids would directly
affect spatially derived QC quantities.

```{r}
#| label: plot-fovs
#| fig-width: 9
#| fig-height: 7

plotCellsFovs(
    spe,
    pointSize=0.3,
    pointAlpha=0.35,
    numberSize=3
)
```

### Question

Do the cell centroids appear correctly positioned inside the FOVs?

This is a useful reminder that **QC begins before calculating a score**.

### Zoom into a set of FOVs

A larger dataset also lets us inspect local FOV structure while keeping the
global tissue context.

```{r}
#| label: zoom-fovs
#| fig-width: 9
#| fig-height: 7

focusFovs <- c(98:100, 113:115, 122:124)

plotZoomFovsMap(
    spe,
    mapPointCol="darkgreen",
    colourBy="Mean.DAPI",
    fovs=focusFovs
)
```

We will reuse the same FOVs for detailed polygon-level visualizations.

## Compute imaging-specific QC metrics

`spatialPerCellQC()` enriches the `SpatialExperiment` with per-cell QC metrics.
Because `readCosmxSXE()` already placed the negative probes in the `NegPrb`
alternative experiment, we can use that structure directly.

```{r}
#| label: spatial-qc

spe <- spatialPerCellQC(
    spe,
    rmZeros=TRUE
)

colnames(colData(spe))
```

Let's focus on the metrics that contribute most directly to the integrated
quality assessment.

```{r}
#| label: inspect-qc-metrics

qcMetrics <- c(
    "log2SignalDensity",
    "Area_um",
    "log2Ctrl_total_ratio",
    "log2AspectRatio",
    "dist_border"
)

summary(as.data.frame(colData(spe)[, qcMetrics]))
```

## Look at metrics one at a time

Before computing an integrated score, inspect the marginal distributions.

```{r}
#| label: metric-hist-signal
#| fig-width: 7
#| fig-height: 4.5

plotMetricHist(
    spe,
    metric="log2SignalDensity"
)
```

```{r}
#| label: metric-hist-area
#| fig-width: 7
#| fig-height: 4.5

plotMetricHist(
    spe,
    metric="Area_um"
)
```

::: {.callout-warning title="Do not jump straight to filtering"}
An extreme value can represent technical failure, unusual morphology, or true
biology. The spatial location of the cell helps distinguish these possibilities.
:::

## Put the metrics back into space

Plotting all ~48k polygons at once is possible, and depends on your machine, 
but a focused set of neighboring FOVs is easier to interpret during the workshop.

```{r}
#| label: make-focus-spe

speFocus <- spe[, spe$fov %in% focusFovs]
dim(speFocus)
```

```{r}
#| label: plot-qc-polygons
#| fig-width: 9
#| fig-height: 7

plotPolygons(
    speFocus,
    colourBy="log2SignalDensity"
)
```

```{r}
#| label: plot-qc-polygons-background
#| fig-width: 9
#| fig-height: 7

plotPolygons(
    speFocus,
    colourBy="log2Ctrl_total_ratio"
)
```

### Discussion

Look for cells or regions with:

- low signal density;
- unusual cell area;
- elevated negative-control signal;
- extreme aspect ratio;
- proximity to FOV borders.

Do these features identify the **same cells**, or different kinds of suspicious
cells?

# From QC metrics to the Quality Score

<!-- ~30 min -->

## Compute the integrated Quality Score

The current development version uses `computeQScore()`.

The function learns how the QC variables combine and stores a per-cell
`QScore` in `colData(spe)`.

```{r}
#| label: compute-qscore

set.seed(1713)

spe <- computeQScore(spe)

summary(spe$QScore)
```

The score ranges from **0 to 1**:

- values closer to **0** indicate lower quality;
- values closer to **1** indicate higher quality.

The current model combines signal density, cell size, control-to-total signal,
and — when relevant — morphology together with FOV-border effects.

## What is the model doing?

Conceptually:

```text
                  log2SignalDensity
                         |
Area_um ----------------+---------------- log2Ctrl_total_ratio
                         |
             morphology / border effect
                         |
                         v
                regularized model
                         |
                         v
                      QScore
```

The important point for the workshop is not to memorize the model formula.

> **The score integrates multiple technical signatures that may be weak or
> ambiguous when inspected independently.**

## Inspect the QScore distribution

```{r}
#| label: qscore-hist
#| fig-width: 7
#| fig-height: 4.5

plotMetricHist(
    spe,
    metric="QScore"
)
```

### Question

Would you choose a filtering threshold before seeing this distribution?

SpaceTrooper does not make the threshold a biological truth.  
The score is continuous and should be inspected in the context of the dataset.

## Map quality back onto the tissue

```{r}
#| label: qscore-space
#| fig-width: 8
#| fig-height: 6

speFocus <- spe[, spe$fov %in% focusFovs]

plotPolygons(
    speFocus,
    colourBy="QScore"
) +
    scale_fill_viridis_c(option="plasma")

# plotPolygons(
#     spe,
#     colourBy="QScore"
# ) +
#     scale_fill_viridis_c(option="plasma")
```

Now compare this map to the individual metrics.

### What does the QScore capture?

Ask whether low-QScore cells coincide with:

- low transcript density;
- large or unusual cell shapes;
- high negative-control signal;
- cells close to problematic FOV borders;
- spatially coherent technical regions.

This is the central interpretative step of the practical.

## Flag cells — but only after inspection

SpaceTrooper provides `computeQScoreFlags()` to turn the continuous score into a
binary flag. The threshold is arbitrary and depends on the experiment, tissue, and platform.

```{r}
#| label: qscore-flags-05

spe <- computeQScoreFlags(
    spe,
    qsThreshold=0.5
)

table(spe$low_QScore)
```

Visualize the result first across the complete experiment. This is where the
spatial structure of low-quality cells becomes apparent, including enrichment
along tissue and FOV borders.

```{r}
#| label: qscore-flags-centroids-05
#| fig-width: 9
#| fig-height: 7

plotCentroids(
    spe,
    colourBy="low_QScore"
)
```

Then zoom to the selected FOVs and inspect the same pattern at polygon level.

```{r}
#| label: qscore-flags-space-05
#| fig-width: 8
#| fig-height: 6

speFocus <- spe[, spe$fov %in% focusFovs]

plotPolygons(
    speFocus,
    colourBy="low_QScore"
) +
    scale_fill_manual(
        values=c(
            "TRUE"="red",
            "FALSE"="#c0c8cf"
        )
    )
```

## Thresholds are decisions, not constants

Try a more stringent threshold:

```{r}
#| label: qscore-flags-075

spe <- computeQScoreFlags(
    spe,
    qsThreshold=0.75
)

table(spe$low_QScore)
```

```{r}
#| label: qscore-flags-space-075
#| fig-width: 8
#| fig-height: 6

speFocus <- spe[, spe$fov %in% focusFovs]

plotPolygons(
    speFocus,
    colourBy="low_QScore"
) +
    scale_fill_manual(
        values=c(
            "TRUE"="red",
            "FALSE"="#c0c8cf"
        )
    )
```

::: {.callout-important title="Interpret before filtering"}
The useful question is not *“What is the correct universal QScore threshold?”*

It is:

**“Which technical artifacts are being captured, and what filtering decision is
appropriate for this experiment?”**
:::

# Exercise: diagnose low-quality cells

<!-- ~15 min -->

Work in small groups or individually.

1. Choose **two individual QC metrics**.
2. Inspect their distributions.
3. Map them spatially.
4. Compare them with `QScore`.
5. Choose a candidate QScore threshold.
6. Explain **which type of artifact** your filtering decision appears to target.

Useful functions:

```r
plotMetricHist()
plotCentroids()
plotPolygons()
computeQScoreFlags()
```

::: {.callout-tip title="Bonus"}
Instead of starting from QScore, select a suspicious spatial region first and
ask which QC variables explain why its cells receive low scores.
:::

# SpaceTrooper: take-home messages

<!-- ~10 min -->

1. **Imaging-based spatial QC has failure modes that are not captured by
   sequencing-derived metrics alone.**

2. `SpatialExperiment` provides the interoperable data infrastructure;
   **SpaceTrooper provides specialized methodology on top of it.**

3. SpaceTrooper combines **expression-derived, morphological, and
   platform-specific information**.

4. `QScore` is a **continuous multivariate quality measure**, not an automatic
   biological truth.

5. QC decisions should always be interpreted **in space**.

# Workshop outlook

We have moved through the complete story:

```text
OSTA / Bioconductor ecosystem
            |
            v
     SpatialExperiment
      /            \
sequencing       imaging
                    |
                    v
               SpaceTrooper
                    |
        +-----------+-----------+
        |           |           |
     metrics     QScore      spatial QC
```

The broader message is:

::: {.callout-note title="Infrastructure first, specialization later"}
A flexible common representation makes it possible to build specialized
technology-aware methods without fragmenting the analysis ecosystem.
:::

# Resources

- **OSTA** — Orchestrating Spatial Transcriptomics Analysis with Bioconductor
- **OSTA.data** — example datasets used throughout OSTA
- **SpatialExperiment** — core spatial experiment infrastructure
- **SpaceTrooper** — imaging-based spatial omics quality control
- **SpaceTrooper preprint** — Banzi et al., *SpaceTrooper: a quality control
  framework for imaging-based spatial omics data*
- **SpaceTrooper development repository** — `drighelli/SpaceTrooper`, branch
  `devel` (same as Bioconductor devel version)

# Session information

```{r}
#| label: session-info

sessionInfo()
```
